The impacts of park access on health outcomes: A spatial comprehensive approach

### 2019-03-04 manuscript.Rnw
Changes suggested by Kari and Nico, plus a couple of others.

\corrauth{Gregory Macfarlane\\
430 Engineering Building\\
Brigham Young University\\
Provo, UT 84602\\
USA}
\begin{abstract}
Recent research has underscored the potential for public green spaces to
influence individual and societal health outcomes, but
empirical measurements of such influences have yielded mixed results to date, with
particular disagreement surrounding how access to parks ought to be defined while
controlling alternate explanations. In this paper we present a comprehensive
measure of park accessibility drawn from random utility choice theory, which avoids
arbitrary assertions of proximity while incorporating potentially numerous
amenities and attributes of both the parks and the population. We apply this
choice-based accessibility measure to correlate Census tract-level obesity and
physical activity rate estimates from the Centers for Disease Control and
Prevention's 500 Cities project with tract-level American Community Survey
socioeconomic data in New York City, paired with geographic parks data from New
York City and geolocated tweet data collected from the Twitter API as a metric of
a park's activity levels. Controlling for the socioeconomic variables and
spatially correlated error terms, we show a positive and significant
relationship between park access and physical activity rates, and a clear and
significant negative relationship between park access and obesity rates.
Applying a more traditional analysis considering 1/4 mile acess to parks yields
mixed results and cannot accommodate features such as park size and twitter
activity.
\end{abstract}

@@

bind_rows(tract_summary, park_summary) %>%
  separate(variable, c("var", "desc"), sep = ": ") %>%
  mutate(
    msd = str_replace(msd, " \\(", "\\\\\\\\ \\("),
    msd = str_c("\\makecell{", msd, "}")
  ) %>%
  kable(escape = FALSE, digits = 2,
        table.envir = "table*", booktabs = TRUE,
        caption = "Descriptive Statistics of Tract and Park Variables",
        col.names = c(" ", "Description", "Minimum", "Median (IQR)", "Maximum", "Source")) %>%
  kable_styling(latex_options = "scale_down") %>%
  column_spec(1, width = "0.5in") %>%
  column_spec(2, width = "2in") %>%
  group_rows(paste("Tract Variables, N =", nrow(tracts)), 1, 14) %>%
  group_rows(paste("Park Variables, N =", nrow(parks)), 15, 16)

@@

Using tracts as the unit of observation in the model allows us to consider the
availability of parks at a sub-metropolitan level, but it also introduces the problem
of spatial correlation of unobserved errors and spatial dependence in the data
generating process \citep{Cliff1970, Anselin1980}. Spatial correlation occurs
when spatially-distributed unobservable or missing variables (school quality,
neighborhood prestige, etc.) influence the modeled outcomes. Spatial dependence
occurs when the outcome {\em depends} on the outcome in neighboring areas; for
example, seeing neighbors exercising may encourage exercise. The relaxed spatial
Durbin model (SDM) proposed by \citet{Burridge1981} controls for both processes:
\begin{equation}\label{eq:sdm}
  \vect{y}= \rho W\vect{y} + X\vect{\beta} + WX\vect{\gamma} + \vect{\epsilon}
\end{equation}
where $\rho$ is the estimated spatial correlation between the outcome variables
$y_i, y_j, \ldots  \in \vect{y}$, $W$ is a matrix specifying the spatial
relationship between observations $i$ and $j$ in the dataset,
$\vect{\beta}$ is a vector of estimable coefficients relating the attributes $X$
of an observation to its outcome, and $\vect{\gamma}$ is a vector of
estimated coefficients relating the attributes of an observation's {\em neighbors}
to its outcome. If only correlation and not dependence exists in the data
generating process, it may  be more efficient to estimate a restricted
spatial autoregressive model and therefore
save degrees of freedom in the estimation. \citet{tra_autoregression_paper}
illustrate a decision algorithm originally proposed by \citet{Florax2003} which uses
a likelihood ratio test of the SDM against the restricted spatial error model (SEM):
\begin{align}
{\boldsymbol y} &= X{\boldsymbol \beta} + {\boldsymbol u},  {\boldsymbol u} = \lambda W {\boldsymbol u} +
{\boldsymbol \epsilon} \label{eq:SEM}\\
{\boldsymbol y} &= X{\boldsymbol \beta} + (I-\lambda W)^{-1}{\boldsymbol \epsilon}\notag
\end{align}
If the likelihood ratio test rejects that the models are equivalent, then the SDM
should be used because its estimates are unbiased. If the test cannot reject
equivalence, the SEM is more efficient.

The researcher must assert $W$ {\em a priori}, though there are several
common specifications \citep{Dubin1998}. In this study we specify that tracts
with population-weighted centroids located within 1.8 Euclidean miles of each
other are
considered neighbors, and the strength of the relationship is the inverse of
the distance between them:
\begin{equation}\label{eq:weightsrule}
W_{ij} = \begin{cases}
  {\frac{1/d_{ij}} {\sum_{k=1}^n( 1/d_{ik})}} & \text{for $d_{ij} \leq 1.8$ miles}\\
  0  & \text{for $d_{ij} > 1.8$ miles}
  \end{cases}
\end{equation}
<<weights, warning = FALSE>>=
# Instead of using the centroid of the tract geometry, we want to use the
# tract's population-weighted centroid.
pop_weighted_centroids <- tracts %>% tbl_df() %>%
  dplyr::select(geoid) %>%
  left_join(
    read_csv("data/CenPop2010_Mean_TR36.csv") %>%
      transmute(geoid = str_c(STATEFP, COUNTYFP, TRACTCE), LATITUDE, LONGITUDE)
  ) %>%
  # convert to sf and project into the feet-based projection used in the
  # parks dataset.
  st_as_sf(crs = 4326, coords = c("LONGITUDE", "LATITUDE")) %>%
  st_transform(2263)

# determine which tracts are neighbors of each other. Need to be less than
# 1.8 miles away from the centroids
tracts.dnn <- dnearneigh(as(pop_weighted_centroids, "Spatial"), 0, 1.8 * 5280)
# calculate the inverse distance to use as a weight on the neighbor connection
dists <- nbdists(tracts.dnn, as(pop_weighted_centroids, "Spatial"))
dists.inv <- lapply(dists, function(x) 1 / x)
# get the spatial weights matrix and its trace to use in any impact analyses
W <- nb2listw(neighbours = tracts.dnn, glist = dists.inv,
              zero.policy = TRUE, style = "W")
trMC <- trW(as(W, "CsparseMatrix"), type="MC") # trace used in montecarlo impacts
@
This is a similar specification to that used by  \citet{tra_autoregression_paper}
in a study of home prices in Atlanta. We selected this specification after
comparing overall model likelihoods between it and specification where tracts
sharing a boundary are considered neighbors (queen-adjacent).

@@

We estimated SDM and SEM models regressing the physical activity rate against
against the base covariates (without the accessibility logsum). A likelihood
ratio test failed to reject that the SDM is different from the SEM, so we use
the SEM specification only going forward. Table \ref{tab:pa_models} presents the
estimated coefficients for the base model with the socioecononomic controls and
no representation of park accessibility. For the most part the coefficients are
highly significant and of the expected sign. Tracts with higher shares
of full-time workers, college-educated adults, young adults, and high-income
households all show a greater share of individuals engaging in regular physical
activity. Conversely, tracts with greater population density and a greater
share of single adults, children, seniors, minorities of all types, and low
income households all have lower modeled rates of physical activity.

To this base model we add a measure of park accessibility considered in two
ways: a logsum-based statistic as described earlier and a more typical quarter-mile
binary access buffer. The estimated coefficients for these models are also
given in Table \ref{tab:pa_models}. In both accessibility specifications,
the access measure is positively correlated with physical activity rates; all else
equal, people in tracts with higher accessibility to parks have higher estimated
rates of physical activity. Though the estimated coefficient on the quarter-mile
buffer appears to have a larger absolute value than the logsum-based statistic,
this masks a key difference in the specification worth further comment.
The logsum allow for variation in access unavailable in the buffer-based method.
In our dataset almost every tract
(\Sexpr{round(sum(tracts$access_025) / nrow(tracts) * 100, 0)} \%) is within
a quarter-mile of {\em some} kind of park, but this park may not be large enough
to support physical activities all individuals would like to participate in.
Because tracts with only adequate park access are grouped with tracts having
excellent park access, the overall effect is comparatively unclear. The logsum-based
coefficient, on the other hand, suggests going from the least-accessible ($-3.0$)
to the most-accessible ($3.0$) tract implies roughly a $1.1$\% percentage-point
rise in estimated physical activity rates

@@

\begin{table*}
  \caption{SEM Coefficients Predicting Physical Activity Rates}
  \label{tab:pa_models}
<<pamodelchart, results="asis">>=
termnames <- c("(Intercept)", "log(density)", "$\\log($Income$)$",
               "Fulltime", "College-educated",  "Single Adults",
               "Youth (0-17)", "Young adults (18-34)",
               "Seniors (65+)", "Black population share",
               "Asian population share",
               "Hispanic population share", "Other Minorities",
               "$\\lambda$", "Access logsum", "1/4 mile access")

texreg(pa_models, digits = 4, table = FALSE,
       booktabs = TRUE, use.packages = FALSE,
       custom.coef.names = termnames)
@
\end{table*}

@@

We now consider the impacts of a model where the dependent variable is
the {\em obesity} rate, and physical activity becomes an independent covariate
alongside the controlling variables and the accessibility metric.
Table \ref{tab:obese_models} again presents the base model without
any accessiblity measure, and models with accessibility captured through a
logsum and through a 1/4 mile buffer. As in the
physical activity models, the coefficients are typically significant and
on the expected sign, with a few exceptions: in this case an increasing hispanic
population share  and an increasing share of children have no significant
relationship on obesity. Unlike in the physical activity models however,
the inference on the accessibility statistic {\em does} change between
specifications. The 1/4 mile buffer neither substantially improves the model
fit nor is its term significant at any threshold, but the accessibility logsum
indicates a significant negative correlation: moving from the least to most
accessible tract decreases the expected obesity rate by approximately half a
percentage point.

\begin{table*}
  \caption{SEM Coefficients Predicting Obesity Rates}
  \label{tab:obese_models}
<<obesemodelchart, results="asis">>=
termnames <- c("(Intercept)", "log(density)", "$\\log($Income$)$",
               "Fulltime", "College-educated",  "Single Adults",
               "Youth (0-17)", "Young adults (18-34)",
               "Seniors (65+)", "Black population share",
               "Asian population share",
               "Hispanic population share", "Other Minorities",
               "Physical activity",
               "$\\lambda$", "Access logsum", "1/4 mile access")
texreg(obesity_models, digits = 4, table = FALSE,
       booktabs = TRUE, use.packages = FALSE,
       custom.coef.names = termnames,  stars = c(0.001, 0.01, 0.05))
@
\end{table*}

>>>>

\corrauth{Gregory Macfarlane\\
430 Engineering Building\\
Brigham Young University\\
Provo, UT 84602\\
USA}
\begin{abstract}
Recent research has underscored the potential for public green spaces to
influence individual and societal health outcomes, but
empirical measurements of such influences have yielded mixed results to date, with
particular disagreement surrounding how access to parks ought to be defined while
controlling alternate explanations. In this paper we present a comprehensive
measure of park accessibility drawn from random utility choice theory, which avoids
arbitrary assertions of proximity while incorporating potentially numerous
amenities and attributes of both the parks and the population. We apply this
choice-based accessibility measure to correlate Census tract-level obesity and
physical activity rate estimates from the Centers for Disease Control and
Prevention's 500 Cities project with tract-level American Community Survey
socioeconomic data in New York City, paired with geographic parks data from New
York City and geolocated tweet data collected from the Twitter API as a metric of
a park's activity levels. Controlling for the socioeconomic variables and
spatially correlated error terms, we show a positive and significant
relationship between park access and physical activity rates, and a clear and
significant negative relationship between park access and obesity rates.
Applying a more traditional analysis considering 1/4 mile access to parks yields
mixed results and cannot accommodate features such as park size and twitter
activity.
\end{abstract}

@@

bind_rows(tract_summary, park_summary) %>%
  separate(variable, c("var", "desc"), sep = ": ") %>%
  mutate(
    msd = str_replace(msd, " \\(", "\\\\\\\\ \\("),
    msd = str_c("\\makecell{", msd, "}")
  ) %>%
  kable(escape = FALSE, digits = 2,
        table.envir = "table*", booktabs = TRUE,
        caption = "Descriptive Statistics of Tract and Park Variables",
        col.names = c(" ", "Description", "Minimum", "Median (IQR)", "Maximum", "Source")) %>%
  kable_styling(latex_options = "scale_down") %>%
  column_spec(2, width = "2in") %>%
  group_rows(paste("Tract Variables, N =", nrow(tracts)), 1, 14) %>%
  group_rows(paste("Park Variables, N =", nrow(parks)), 15, 16)

@@

Using tracts as the unit of observation in the model allows us to consider the
availability of parks at a sub-metropolitan level, but it also introduces the problem
of spatial correlation of unobserved errors and spatial dependence in the data
generating process \citep{Cliff1970, Anselin1980}. Spatial correlation occurs
when spatially-distributed unobservable or missing variables (school quality,
neighborhood prestige, etc.) influence the modeled outcomes. Spatial dependence
occurs when the outcome {\em depends} on the outcome in neighboring areas; for
example, seeing neighbors exercising may encourage exercise. The relaxed spatial
Durbin model (SDM) proposed by \citet{Burridge1981} controls for both processes:
\begin{equation}\label{eq:sdm}
  \vect{y}= \rho W\vect{y} + X\vect{\beta} + WX\vect{\gamma} + \vect{\epsilon}
\end{equation}
where $\rho$ is the estimated spatial correlation between the outcome variables
$y_i, y_j, \ldots  \in \vect{y}$, $W$ is a matrix specifying the spatial
relationship between observations $i$ and $j$ in the dataset,
$\vect{\beta}$ is a vector of estimable coefficients relating the attributes $X$
of an observation to its outcome, and $\vect{\gamma}$ is a vector of
estimated coefficients relating the attributes of an observation's {\em neighbors}
to its outcome. If only correlation and not dependence exists in the data
generating process, it may  be more efficient to estimate a restricted
spatial autoregressive model and therefore
save degrees of freedom in the estimation. \citet{tra_autoregression_paper}
illustrate a decision algorithm originally proposed by \citet{Florax2003} which uses
a likelihood ratio test of the SDM against the restricted spatial error model (SEM):
\begin{align}
{\boldsymbol y} &= X{\boldsymbol \beta} + {\boldsymbol u},  {\boldsymbol u} = \gamma W {\boldsymbol u} +
{\boldsymbol \epsilon} \label{eq:SEM}\\
{\boldsymbol y} &= X{\boldsymbol \beta} + (I-\gamma W)^{-1}{\boldsymbol \epsilon}\notag
\end{align}
If the likelihood ratio test rejects that the models are equivalent, then the SDM
should be used because its estimates are unbiased. If the test cannot reject
equivalence, the SEM is more efficient.

The researcher must assert $W$ {\em a priori}, though there are several
common specifications \citep{Dubin1998}. In this study we specify that tracts
with population-weighted centroids located within 1.8 Euclidean miles of each
other are considered neighbors, and the strength of the relationship is the
row-standardized (each row sums to 1) inverse of the distance between them:
\begin{equation}\label{eq:weightsrule}
W_{ij} = \begin{cases}
  {\frac{d_{ij}^{-1}} {\sum_{k=1}^n d_{ik}^{-1}}} & \text{for $d_{ij} \leq 1.8$ miles}\\
  0  & \text{for $d_{ij} > 1.8$ miles}
  \end{cases}
\end{equation}
<<weights, warning = FALSE>>=
# Instead of using the centroid of the tract geometry, we want to use the
# tract's population-weighted centroid.
pop_weighted_centroids <- tracts %>% tbl_df() %>%
  dplyr::select(geoid) %>%
  left_join(
    read_csv("data/CenPop2010_Mean_TR36.csv") %>%
      transmute(geoid = str_c(STATEFP, COUNTYFP, TRACTCE), LATITUDE, LONGITUDE)
  ) %>%
  # convert to sf and project into the feet-based projection used in the
  # parks dataset.
  st_as_sf(crs = 4326, coords = c("LONGITUDE", "LATITUDE")) %>%
  st_transform(2263)

# determine which tracts are neighbors of each other. Need to be less than
# 1.8 miles away from the centroids
tracts.dnn <- dnearneigh(as(pop_weighted_centroids, "Spatial"), 0, 1.8 * 5280)
# calculate the inverse distance to use as a weight on the neighbor connection
dists <- nbdists(tracts.dnn, as(pop_weighted_centroids, "Spatial"))
dists.inv <- lapply(dists, function(x) 1 / x)
# get the spatial weights matrix and its trace to use in any impact analyses
W <- nb2listw(neighbours = tracts.dnn, glist = dists.inv,
              zero.policy = TRUE, style = "W")
trMC <- trW(as(W, "CsparseMatrix"), type="MC") # trace used in montecarlo impacts
@
This is a similar specification to that used by  \citet{tra_autoregression_paper}
in a study of home prices in Atlanta. We selected this specification after
comparing overall model likelihoods between it and a specification where tracts
sharing a boundary are considered neighbors.

@@

We estimated SDM and SEM models regressing the physical activity rate
against the base covariates (without the accessibility logsum). A likelihood
ratio test failed to reject that the SDM is different from the SEM, so we use
the SEM specification only going forward. Table \ref{tab:pa_models} presents the
estimated coefficients for the base model with the socioecononomic controls and
no representation of park accessibility. For the most part the coefficients are
highly significant and of the expected sign. Tracts with higher shares
of full-time workers, college-educated adults, young adults, and high-income
households all show a greater share of individuals engaging in regular physical
activity. Conversely, tracts with greater population density and a greater
share of single adults, children, seniors, minorities of all types, and low
income households all have lower modeled rates of physical activity.

To this base model we add a measure of park accessibility considered in two
ways: a logsum-based statistic as described earlier and a more typical quarter-mile
binary access buffer. The estimated coefficients for these models are also
given in Table \ref{tab:pa_models}. In both accessibility specifications,
the access measure is positively correlated with physical activity rates; all else
equal, people in tracts with higher accessibility to parks have higher estimated
rates of physical activity. Though the estimated coefficient on the quarter-mile
buffer appears to have a larger absolute value than the logsum-based statistic,
this masks a key difference in the specification worth further comment.
The logsum allows for variation in access unavailable in the buffer-based method.
In our dataset almost every tract
(\Sexpr{round(sum(tracts$access_025) / nrow(tracts) * 100, 0)} \%) is within
a quarter-mile of {\em some} kind of park, but this park may not be large enough
to support physical activities all individuals would like to participate in.
Because tracts with only adequate park access are grouped with tracts having
excellent park access, the overall effect is comparatively unclear. The logsum-based
coefficient, on the other hand, suggests going from the least-accessible ($-3.0$)
to the most-accessible ($3.0$) tract implies roughly a $1.1$ percentage point
rise in estimated physical activity rates.

@@

\begin{table*}
  \caption{SEM Coefficients Predicting Physical Activity Rates}
  \label{tab:pa_models}
<<pamodelchart, results="asis">>=
termnames <- c("(Intercept)", "log(density)", "$\\log($Income$)$",
               "Fulltime", "College-educated",  "Single Adults",
               "Youth (0-17)", "Young adults (18-34)",
               "Seniors (65+)", "Black population share",
               "Asian population share",
               "Hispanic population share", "Other Minorities",
               "$\\gamma$: spatial correlation", "Access logsum", "1/4 mile access")

texreg(pa_models, digits = 4, table = FALSE,
       booktabs = TRUE, use.packages = FALSE,
       custom.coef.names = termnames,
       custom.note = "%stars. Standard errors in parentheses.")
@
\end{table*}

@@

We now consider the impacts of a model where the dependent variable is
the {\em obesity} rate, and physical activity becomes an independent covariate
alongside the controlling variables and the accessibility metric.
Table \ref{tab:obese_models} again presents the base model without
any accessiblity measure, and models with accessibility captured through a
logsum and through a 1/4 mile buffer. As in the
physical activity models, the coefficients are typically significant and
of the expected sign, with a few exceptions: in this case an increasing hispanic
population share  and an increasing share of children have no significant
relationship on obesity. Unlike in the physical activity models however,
the inference on the accessibility statistic {\em does} change between
specifications. The 1/4 mile buffer neither substantially improves the model
fit nor is its term significant at any threshold, but the accessibility logsum
indicates a significant negative correlation: moving from the least to most
accessible tract decreases the expected obesity rate by approximately half a
percentage point.

\begin{table*}
  \caption{SEM Coefficients Predicting Obesity Rates}
  \label{tab:obese_models}
<<obesemodelchart, results="asis">>=
termnames <- c("(Intercept)", "log(density)", "$\\log($Income$)$",
               "Fulltime", "College-educated",  "Single Adults",
               "Youth (0-17)", "Young adults (18-34)",
               "Seniors (65+)", "Black population share",
               "Asian population share",
               "Hispanic population share", "Other Minorities",
               "Physical activity",
               "$\\gamma$: spatial correlation", "Access logsum", "1/4 mile access")
texreg(obesity_models, digits = 4, table = FALSE,
       booktabs = TRUE, use.packages = FALSE,
       custom.coef.names = termnames,
       custom.note = "%stars. Standard errors in parentheses.")
@
\end{table*}
